On Adding Pattern Matching to Haskell-based Deeply Embedded Domain Specific Languages

### 2020-11-25 paper/paper.tex
Add more plugin explanation; remove some blank lines in code snippets

\begin{lstlisting}
data E t where
  ...
  ConstructRep :: (Typeable a, ERep a) =>
    E (ERepTy a) -> E a

  LeftExp :: E a -> E (Either a b)
  RightExp :: E b -> E (Either a b)

  PairExp :: E a -> E b -> E (a, b)
  ...
\end{lstlisting}

@@

Within the \ttt{E} expression language, a pattern match is represented by the
\ttt{CaseExp} constructor:

\begin{lstlisting}
data E t where
  ...
  CaseExp :: (ERep t, ERepTy (ERepTy t) ~ ERepTy t) =>
    E t -> E (SumMatch (ERepTy t) r) -> E r
  ...
\end{lstlisting}

@@

% \newpage
\begin{lstlisting}
data E t where
  ...
  SumMatchExp :: (ERep a, ERep b, ERepTy b ~ b) =>
    E (ProdMatch a r) -> E (SumMatch b r)
      -> E (SumMatch (Either a b) r)
  OneSumMatch ::
    (ERep a, ERep b, ERepTy a ~ a) =>
      E (ProdMatch a b) -> E (SumMatch a b)
  EmptyMatch :: (ERep b) => E (SumMatch Void b)
\end{lstlisting}

@@

\begin{lstlisting}
data E t where
  ...
  StepExp :: E b -> E (Iter a b)
  DoneExp :: E a -> E (Iter a b)

  TailRec :: (ERep a, ERep b) =>
    E (b -> Iter a b) -> E (b -> a)
  ...
\end{lstlisting}

@@

envLookup :: Typeable a => Env -> Name a -> Maybe (E a)
envLookup (Env maps) = go maps
  where
    go []                = Nothing

    go ((n' :=> e):rest) =
      case namesEq n n' of
        Just Refl -> Just e
        Nothing   -> go rest
\end{lstlisting}

@@

sumMatchAbs :: (ERepTy (ERepTy s) ~ ERepTy s, ERep s)
  => Env -> E (SumMatch (ERepTy s) t) -> s -> t
...

prodMatchAbs :: (ERep s) =>
  Env -> E (ProdMatch s t) -> s -> t
...
\end{lstlisting}

@@

For example, \ttt{externalize x} marks the expression \ttt{x}.

If an expression already has an EDSL type (a type of the form \ttt{E a} for some
\ttt{a}), then the marking procedure ignores it and does not wrap it with
\ttt{externalize} (see $M$ in Figure~\ref{fig:Rewrites}).

The plugin's control flow is described in Figure~\ref{fig:CorePlugin}. When control
flow reaches a node, the rewrite rules listed in parentheses are performed. These rewrite
rule names correspond to the rules given in Figure~\ref{fig:Rewrites}.

% In the flowchart given in Figure~\ref{fig:CorePlugin}, the names in parentheses
% in a node refer to the names of transformations given in the rewrite rules in
% Figure~\ref{fig:Rewrites}. The control flow of the plugin is described by the
% flowchart and the rewrites performed when control reaches a node is given by the list of
% rule names in parentheses in that node.

In Figure~\ref{fig:Rewrites},
$\varepsilon$ represents an empty term (in the context of the term rewriting system used in that Figure). Calls
to \ttt{unrep :: E a -> a} are only used internally and will not exist in
the final result of the transformation.

>>>>

\begin{lstlisting}
data E t where
  ...
  ConstructRep :: (Typeable a, ERep a) =>
    E (ERepTy a) -> E a
  LeftExp :: E a -> E (Either a b)
  RightExp :: E b -> E (Either a b)
  PairExp :: E a -> E b -> E (a, b)
  ...
\end{lstlisting}

@@

Within the \ttt{E} expression language, a pattern match is represented by the
\ttt{CaseExp} constructor:
\begin{lstlisting}
data E t where
  ...
  CaseExp :: (ERep t, ERepTy (ERepTy t) ~ ERepTy t) =>
    E t -> E (SumMatch (ERepTy t) r) -> E r
  ...
\end{lstlisting}

@@

% \newpage
\begin{lstlisting}
data E t where
  ...
  SumMatchExp :: (ERep a, ERep b, ERepTy b ~ b) =>
    E (ProdMatch a r) -> E (SumMatch b r)
      -> E (SumMatch (Either a b) r)

  OneSumMatch ::
    (ERep a, ERep b, ERepTy a ~ a) =>
      E (ProdMatch a b) -> E (SumMatch a b)
  EmptyMatch :: (ERep b) => E (SumMatch Void b)
\end{lstlisting}

@@

\begin{lstlisting}
data E t where
  ...
  StepExp :: E b -> E (Iter a b)
  DoneExp :: E a -> E (Iter a b)
  TailRec :: (ERep a, ERep b) =>
    E (b -> Iter a b) -> E (b -> a)
  ...
\end{lstlisting}

@@

envLookup :: Typeable a => Env -> Name a -> Maybe (E a)
envLookup (Env maps) = go maps
  where
    go []                = Nothing
    go ((n' :=> e):rest) =
      case namesEq n n' of
        Just Refl -> Just e
        Nothing   -> go rest
\end{lstlisting}

@@

sumMatchAbs :: (ERepTy (ERepTy s) ~ ERepTy s, ERep s)
  => Env -> E (SumMatch (ERepTy s) t) -> s -> t
...
prodMatchAbs :: (ERep s) =>
  Env -> E (ProdMatch s t) -> s -> t
...
\end{lstlisting}

@@

For example, \ttt{externalize x} marks the expression \ttt{x}.  If an expression
already has an EDSL type (a type of the form \ttt{E a} for some \ttt{a}), then
the marking procedure ignores it and does not wrap it with \ttt{externalize}
(see $M$ in Figure~\ref{fig:Rewrites}).

These \verb|externalize| marks are pushed through the input program, starting at
the outermost level. The plugin repeatedly runs over each newly marked node in
the abstract syntax tree, applying the appropriate rewrite rules as it goes.
This process pushes the marks through the program until it reaches a base case
that has no proper subexpressions, such as a character literal.

The plugin's control flow is described in Figure~\ref{fig:CorePlugin}. When control
flow reaches a node, the rewrite rules listed in parentheses are performed. These rewrite
rule names correspond to the rules given in Figure~\ref{fig:Rewrites}.
% In the flowchart given in Figure~\ref{fig:CorePlugin}, the names in parentheses
% in a node refer to the names of transformations given in the rewrite rules in
% Figure~\ref{fig:Rewrites}. The control flow of the plugin is described by the
% flowchart and the rewrites performed when control reaches a node is given by the list of
% rule names in parentheses in that node.
In Figure~\ref{fig:Rewrites},
$\varepsilon$ represents an empty term (in the context of the term rewriting system used in that Figure). Calls
to \ttt{unrep :: E a -> a} are only used internally and will not exist in
the final result of the transformation.
